---
title: "Your Title"
subtitle: "BMIN 5030 Final Project"
author: "Lee Krobock"
format: html
editor: visual
number-sections: true
embed-resources: true
---

------------------------------------------------------------------------

## Overview {#sec-overview}

A relatively recent development in preventative medicine, the human papillomavirus (HPV) vaccine confers immunity to multiple oncogenic strains of HPV.  Studies have demonstrated decreased risks of head, neck, and cervical cancers, and cervical dysplasia in vaccinated individuals.^1^  Despite this, HPV vaccination has proven to be a challenging task in practice, with multiple changes to vaccine recommendations since its initial release and notable vaccine hesitancy, particularly early in its history and among parents of children and adolescents.  The goal of this project is to analyze questionnaire data from the National Health and Nutrition Examination Survey (NHANES) from 2013 to 2023 to describe trends in HPV vaccine coverage and vaccination practices in the United States.

[Final Project GitHub repository](https://github.com/lkrobock/BMIN-5030-Final-Project)

## Introduction {#sec-introduction}

The first vaccine against human papillomavirus, Gardasil 4vHPV, was licensed for use by the U.S. Food and Drug Administration (FDA) in June 2006, followed by official recommendations for use by the Advisory Committee on Immunization Practices (ACIP) published in March 2007.^2^ This quadrivalent vaccine provided protection against four strains of HPV and was initially recommended only for female patients between the ages of 9 and 26 years.^2^ Subsequently, in 2010, these recommendations were expanded to include males between ages 9 and 26.^3^ Following the 2015 FDA approval of a nonavalent HPV vaccine, which provided protection against an additional five disease-associated strains, the ACIP updated its recommendations to consider two appropriately-timed doses of HPV vaccine to be a completed series in patients who received the first dose between ages 9 and 14.^4^ The most recent update to HPV immunization guidelines recommends “shared clinical decision-making” to guide vaccination in adults between ages 27 and 45 years old, and notes that HPV vaccines are not licensed for use in adults over age 45.^5^

There are significant and well-described benefits to HPV vaccination– studies have demonstrated decreased risk of strain-specific HPV infection and associated growths, dysplasias, and cancers.^1,6^ Furthermore, Ellingson et al. noted in their systematic review that multiple studies suggest increased vaccine effectiveness with younger patient age at vaccination.^6^ In practice, however, clinicians have encountered challenges in implementing HPV vaccination as a standard preventative care practice.  In addition to slow shifts in practice following official recommendations and guideline updates, providers describe patient hesitancy early in the history of the HPV vaccine, related to potential side effects, adverse effects, and concerns about children and adolescents becoming sexually active earlier following vaccination.  In one clinician’s experience, it was not until the link between HPV and cancer became more widely understood by the public and HPV vaccination was marketed toward adults in addition to pediatric patients that it gained traction with the public.

Preventative healthcare, including immunization, is an interdisciplinary pursuit that requires insight and participation from clinicians, health educators, epidemiologists and public health researchers, medical and industry researchers, policymakers, and patient stakeholders.  By analyzing the data, I hope to identify and describe changes in HPV vaccination practices that may inform ongoing efforts in health promotion and preventative care.

## Methods {#sec-methods}

The National Health and Nutrition Examination Survey, or NHANES, is a multimodal study program initially conducted by the Centers for Disease Control (CDC) in the early 1960s, intended to provide nationally-representative data on health and nutrition in the United States. Over the past 25 years, the NHANES program evolved into an annual study consisting of health and dietary questionnaires, physical examinations, and laboratory testing. Data is typically reported in two year cycles, during which approximately 5,000 subjects are examined. Sampling and weighting procedures are designed and reviewed with each data cycle so the NHANES captures information that can be considered representative of the entire United States.

The NHANES program was suspended from March 2020 to August 2021 due to the COVID-19 pandemic. Its resumption in 2021 was marked by changes in sampling and study design to mitigate the ongoing spread of COVID-19, and decreased response rates consistent with trends described among other large scale U.S. surveys and studies.^7^ The most recently published NHANES datasets available to the public are from this first post-pandemic survey cycle, conducted from 2021 to 2023.

In 2024, the nhanesA R package was published, providing a means to directly import, search, and review data stored on the CDC's NHANES website in RStudio. Otherwise, NHANES data is stored digitally in SAS transport format files and can be viewed and converted to other formats using the SAS Universal Viewer program.^8^

### Loading required R packages

```{r}

library(tidyverse)  #dplyr, readr, ggplot2, stringr + tibble, tidyr
library(nhanesA)
library(gtsummary)
library(cowplot)
library(paletteer)
library(viridis)

```

### Loading NHANES demographics datasets

```{r}

nhanes2013d <- nhanes('DEMO_H')
nhanes2015d <- nhanes('DEMO_I')
nhanes2017d <- nhanes('P_DEMO') 

# Note: Naming convention for 2017-2020 combined, nationally representative pre-pandemic data 'P_filename' differs from the naming format of other data sets.

nhanes2021d <- nhanes('DEMO_L')

```

### Demographics data cleaning and processing

```{r}

nhanes2013demo <- mutate(nhanes2013d, DataCycle = "2013-2014",
  DataCycle = factor(DataCycle)) |>
  rename(ID = SEQN, Gender = RIAGENDR, Age = RIDAGEYR) |> 
  mutate(nhanes2013d, EligibleSubject = case_when(
    Age >= 9 & Age <= 59 ~ "Eligible",
    Age < 9 | Age > 59 ~ "Ineligible"), 
  EligibleSubject = factor(EligibleSubject)) |>
  select(DataCycle, ID, Gender, Age, EligibleSubject)

# Added a DataCycle variable to identify NHANES cycle year within the dataframe and an EligibleSubject variable to identify subjects in the eligible population sample for the immunization questionnaire.  
# Renamed variables and selected only variables of interest plus subject ID

nhanes2015demo <- mutate(nhanes2015d, DataCycle = "2015-2016",
  DataCycle = factor(DataCycle)) |>
  rename(ID = SEQN, Gender = RIAGENDR, Age = RIDAGEYR) |> 
  mutate(nhanes2015d, EligibleSubject = case_when(
    Age >= 9 & Age <= 59 ~ "Eligible",
    Age < 9 | Age > 59 ~ "Ineligible"), 
  EligibleSubject = factor(EligibleSubject)) |>
  select(DataCycle, ID, Gender, Age, EligibleSubject)

nhanes2017demo <- mutate(nhanes2017d, DataCycle = "2017-2020",
  DataCycle = factor(DataCycle)) |>
  rename(ID = SEQN, Gender = RIAGENDR, Age = RIDAGEYR) |> 
  mutate(nhanes2017d, EligibleSubject = case_when(
    Age >= 9 & Age <= 59 ~ "Eligible",
    Age < 9 | Age > 59 ~ "Ineligible"), 
  EligibleSubject = factor(EligibleSubject)) |>
  select(DataCycle, ID, Gender, Age, EligibleSubject)

# Note: Immunization questionnaire eligiblity changed in the most recent NHANES cycle-- only subjects between ages 9 and 49 were asked about HPV vaccination.  The Results section will examine findings from the 49-59 year old age group in previous cycles.

nhanes2021demo <- mutate(nhanes2021d, DataCycle = "2021-2023",
  DataCycle = factor(DataCycle)) |>
  rename(ID = SEQN, Gender = RIAGENDR, Age = RIDAGEYR) |> 
  mutate(nhanes2021d, EligibleSubject = case_when(
    Age >= 9 & Age <= 49 ~ "Eligible",
    Age < 9 | Age > 49 ~ "Ineligible"), 
  EligibleSubject = factor(EligibleSubject)) |>
  select(DataCycle, ID, Gender, Age, EligibleSubject)

nhanes_demos <- rbind(nhanes2013demo, nhanes2015demo, nhanes2017demo, nhanes2021demo)

# Created one dataframe to contain the selected demographic data from all four data cycles.

```

### Loading NHANES immunization questionnaire datasets

```{r}

nhanes2013i <- nhanes('IMQ_H')
nhanes2015i <- nhanes('IMQ_I')
nhanes2017i <- nhanes('P_IMQ')

# Note: Again, naming convention 'P_filename' for 2017-2020 combined, nationally representative pre-pandemic data differs from the naming format of other data sets.

nhanes2021i <- nhanes('IMQ_L')

```

### Immunization questionnaire data cleaning and processing

```{r}

nhanes2013imm <- nhanes2013i |>
  rename(ID = SEQN,
         AnyHPVFemales = IMQ040,
         AnyHPVMales = IMQ070,
         FirstDoseAge = IMQ090,
         TotalShots = IMQ045) |>
  mutate(AnyHPVShots = case_when(
  (AnyHPVFemales == "Yes" | AnyHPVMales == "Yes") ~ "Yes",
  (AnyHPVFemales == "No" | AnyHPVMales == "No") ~ "No"),
  AnyHPVShots = factor(AnyHPVShots)) |>
  mutate(HPVComplete = case_when((FirstDoseAge >= 9 & TotalShots == "3 Doses") |
  (FirstDoseAge >= 9 & FirstDoseAge <= 14 & TotalShots == "2 Doses") ~ "Complete",
  (FirstDoseAge > 14 & FirstDoseAge <= 59 & TotalShots == "2 Doses") |
  (FirstDoseAge >= 9 | FirstDoseAge == 999  & (TotalShots == "1 Dose" | TotalShots == "Don't know" | TotalShots == "Refused")) ~ "Incomplete"),
  HPVComplete = factor(HPVComplete)) |>
  select(ID, AnyHPVFemales, AnyHPVMales, AnyHPVShots, FirstDoseAge, TotalShots, HPVComplete) 

# Created a "combined" variable for any HPV vaccines received across sex groups.  
# Additionally created a variable using age at first shot and number of shots received to determine if subject has received a complete HPV vaccine series by current standard of 3 shots or 2 shots if first received before age 15.  Unknown or refused total number of shots recoded to incomplete series.
# Renamed variables and selected only variables of interest plus subject ID.

# Note: IMQ questionnaire items are the same but with different IDs in the 2013-2014 dataset.

nhanes2015imm <- nhanes2015i |> 
  rename(ID = SEQN, 
         AnyHPVFemales = IMQ060, 
         AnyHPVMales = IMQ070, 
         FirstDoseAge = IMQ090,
         TotalShots = IMQ100) |>
  mutate(AnyHPVShots = case_when(
  (AnyHPVFemales == "Yes" | AnyHPVMales == "Yes") ~ "Yes",
  (AnyHPVFemales == "No" | AnyHPVMales == "No") ~ "No"),
  AnyHPVShots = factor(AnyHPVShots)) |>
  mutate(HPVComplete = case_when((FirstDoseAge >= 9 & TotalShots == "3 Doses") |
  (FirstDoseAge >= 9 & FirstDoseAge <= 14 & TotalShots == "2 Doses") ~ "Complete", 
  (FirstDoseAge > 14 & FirstDoseAge <= 59 & TotalShots == "2 Doses") |
  (FirstDoseAge >= 9 | FirstDoseAge == 999  & (TotalShots == "1 Dose" | TotalShots == "Don't know" | TotalShots == "Refused")) ~ "Incomplete"),
  HPVComplete = factor(HPVComplete)) |>
  select(ID, AnyHPVFemales, AnyHPVMales, AnyHPVShots, FirstDoseAge, TotalShots, HPVComplete) 

nhanes2017imm <- nhanes2017i |>
  rename(ID = SEQN, 
         AnyHPVFemales = IMQ060, 
         AnyHPVMales = IMQ070, 
         FirstDoseAge = IMQ090,
         TotalShots = IMQ100) |>
  mutate(AnyHPVShots = case_when(
  (AnyHPVFemales == "Yes" | AnyHPVMales == "Yes") ~ "Yes",
  (AnyHPVFemales == "No" | AnyHPVMales == "No") ~ "No"),
  AnyHPVShots = factor(AnyHPVShots)) |>
  mutate(HPVComplete = case_when((FirstDoseAge >= 9 & TotalShots == "3 Doses") |
  (FirstDoseAge >= 9 & FirstDoseAge <= 14 & TotalShots == "2 Doses") ~ "Complete",
  (FirstDoseAge > 14 & FirstDoseAge <= 59 & TotalShots == "2 Doses") |
  (FirstDoseAge >= 9 | FirstDoseAge == 999  & (TotalShots == "1 Dose" | TotalShots == "Don't know" | TotalShots == "Refused")) ~ "Incomplete"),
  HPVComplete = factor(HPVComplete)) |>
  select(ID, AnyHPVFemales, AnyHPVMales, AnyHPVShots, FirstDoseAge, TotalShots, HPVComplete) 

nhanes2021imm <- nhanes2021i |>
  rename(ID = SEQN, 
         AnyHPVFemales = IMQ060, 
         AnyHPVMales = IMQ070, 
         FirstDoseAge = IMQ090,
         TotalShots = IMQ100) |>
  mutate(AnyHPVShots = case_when(
  (AnyHPVFemales == "Yes" | AnyHPVMales == "Yes") ~ "Yes",
  (AnyHPVFemales == "No" | AnyHPVMales == "No") ~ "No"),
  AnyHPVShots = factor(AnyHPVShots)) |>
  mutate(HPVComplete = case_when((FirstDoseAge >= 9 & TotalShots == "3 Doses") |
  (FirstDoseAge >= 9 & FirstDoseAge <= 14 & TotalShots == "2 Doses") ~ "Complete", 
  (FirstDoseAge > 14 & FirstDoseAge <= 49 & TotalShots == "2 Doses") |
  (FirstDoseAge >= 9 | FirstDoseAge == 999  & (TotalShots == "1 Dose" | TotalShots == "Don't know" | TotalShots == "Refused")) ~ "Incomplete"),
  HPVComplete = factor(HPVComplete)) |>
  select(ID, AnyHPVFemales, AnyHPVMales, AnyHPVShots, FirstDoseAge, TotalShots, HPVComplete)

# Created one dataframe to contain the selected demographic data from all four data cycles.
nhanes_imms <- rbind(nhanes2013imm, nhanes2015imm, nhanes2017imm, nhanes2021imm)

```

### Creating joined demographics and immunization datasets

```{r}
# Created combined demographics and immunization questionnaire datasets for each NHANES cycle, and one dataset containing all data cycles.

nhanes2013 <- left_join(nhanes2013demo, nhanes2013imm, by = c("ID"))
nhanes2015 <- left_join(nhanes2015demo, nhanes2015imm, by = c("ID"))
nhanes2017 <- left_join(nhanes2017demo, nhanes2017imm, by = c("ID"))
nhanes2021 <- left_join(nhanes2021demo, nhanes2021imm, by = c("ID"))

nhanes_all <- rbind(nhanes2013, nhanes2015, nhanes2017, nhanes2021)

# Created a dataset specifically for subjects meeting HPV vaccine questionnaire eligibility, i.e. age 9-59 in pre-pandemic datasets and age 9-49 in post-pandemic dataset.

nhanes_hpv_eligible <- filter(nhanes_all, EligibleSubject == "Eligible")

```

*Describe the data used and general methodological approach used to address the problem described in the @sec-introduction. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why.*

## Results {#sec-results}

### Reviewing participant demographics

```{r}

demos_summary <- nhanes_demos |>
  select(DataCycle, Gender, Age, EligibleSubject) |>
  tbl_summary(by = DataCycle) |>
  modify_caption("**Table 1: NHANES Subject Demographics**") |>
  bold_labels()
print(demos_summary)

hpvdemos_summary <- nhanes_hpv_eligible |>
  select(DataCycle, Gender, Age) |>
  tbl_summary(by = DataCycle) |>
  modify_caption("**Table 2: Demographics of HPV Vaccine Questionnaire Subjects**") |>
  bold_labels()
print(hpvdemos_summary)

```

*Comment on general demographics.*

#### Consideration #1:  Subjects between ages 49 and 59 who reported HPV vaccination

```{r}
cutoffsubjects <- filter(nhanes_all, Age > 49 & Age < 59 & (AnyHPVFemales == "Yes" | AnyHPVMales == "Yes"))

count(cutoffsubjects, DataCycle)
count(nhanes_hpv_eligible, DataCycle)
```

-   *NHANES 2013-2014: 14/6019 = 0.00233*

-   *NHANES 2015-2016: 8/5780 = 0.00138*

-   *NHANES 2017-2020: 31/8855 = 0.00350*

*In the first three NHANES data cycles analyzed in this project, subjects ages 49-59 who reported previous HPV vaccination only comprised between 0.1% and 0.3% of the total eligible subject population. This age group is excluded from HPV vaccination questionnaire data in the 2021-2023 NHANES.*

#### Consideration #2:  Subjects reporting HPV vaccination prior to age 9

```{r}
earlyhpvsubjects <- filter(nhanes_all, FirstDoseAge < 9)
earlyhpvsubjects

count(earlyhpvsubjects, DataCycle)
count(nhanes_hpv_eligible, DataCycle)
```

-   *NHANES 2013-2014: 12/6019 = 0.00199*

-   *NHANES 2015-2016: 20/5780 = 0.00346*

-   *NHANES 2017-2020: 42/8855 = 0.00474*

-   *NHANES 2021-2023: 42/5347 = 0.00785*

*While another relatively small group– comprising between approximately 0.1% and 0.7% of the total eligible subject population for each NHANES cycle– individuals reporting initial HPV vaccination before age 9 pose a challenge to data analysis. There is very little information available about HPV immunization in children under 9, and although earlier immunization is generally preferable, current guidelines indicate age 9 as the earliest approved age for HPV vaccination.*

*It is possible that some immunization questionnaire responses were incorrect– especially in cases where the first HPV dose age was reported as age 1 or 2, as hepatitis B (HBV) vaccination is typically completed in the first year of life. Although small and potentially more reflective of data inaccuracy, the proportion of early-vaccinated HPV subjects continuously increased over the past four NHANES cycles, and is worth further monitoring in future years.*

### Question #1: Any HPV Vaccination

```{r}

q1summary <- nhanes_hpv_eligible |>
  select(DataCycle, Gender, AnyHPVFemales, AnyHPVMales) |>
  tbl_summary(by = DataCycle) |>
  remove_row_type(variables = c(AnyHPVFemales, AnyHPVMales), 
                  type = "missing") |>
  modify_caption("**Question #1: Have you ever received one or more doses of HPV vaccine?**") |>
  bold_labels()
q1summary

hpvq1fem1 <- ggplot(data=subset(nhanes_hpv_eligible, !is.na(AnyHPVFemales)), 
  aes(x = DataCycle, fill = AnyHPVFemales)) + 
  geom_bar() + 
  labs(title = "HPV Vaccination, Female (Totals)") +
  labs(x = "NHANES Data Cycle", y = "Number of subjects") +
  theme(title = element_text(size = 10), 
        axis.text.x = element_text(angle = 45, size= 7),
        legend.title = element_text(size = 7), 
        legend.text = element_text(size = 7)) +
  scale_fill_paletteer_d("ggthemes::excel_Celestial")

hpvq1fem2 <- ggplot(data=subset(nhanes_hpv_eligible, !is.na(AnyHPVFemales)), 
  aes(x = DataCycle, fill = AnyHPVFemales)) + 
  geom_bar(position = "fill") + 
  labs(title = "HPV Vaccination, Female (Proportions)") +
  labs(x = "NHANES Data Cycle", y = "Percentage of subjects") +
  theme(title = element_text(size = 10), 
        axis.text.x = element_text(angle = 45, size= 7),
        legend.title = element_text(size = 7), 
        legend.text = element_text(size = 7)) +
  scale_fill_paletteer_d("vangogh::Chaise")

plot_grid(hpvq1fem1, hpvq1fem2, label_size = 3)

hpvq1male1 <- ggplot(data=subset(nhanes_hpv_eligible, !is.na(AnyHPVMales)), 
  aes(x = DataCycle, fill = AnyHPVMales)) + 
  geom_bar() +
  labs(title = "HPV Vaccination, Male (Totals)") +
  labs(x = "NHANES Data Cycle", y = "Number of subjects") +
  theme(title = element_text(size = 10), 
        axis.text.x = element_text(angle = 45, size= 7),
        legend.title = element_text(size = 7), 
        legend.text = element_text(size = 7)) +
  scale_fill_paletteer_d("MetBrewer::Archambault")

hpvq1male2 <- ggplot(data=subset(nhanes_hpv_eligible, !is.na(AnyHPVMales)), 
  aes(x = DataCycle, fill = AnyHPVMales)) + 
  geom_bar(position = "fill") +
  labs(title = "HPV Vaccination, Male (Proportions)") +
  labs(x = "NHANES Data Cycle", y = "Percentage of subjects") +
  theme(title = element_text(size = 10), 
        axis.text.x = element_text(angle = 45, size= 7),
        legend.title = element_text(size = 7), 
        legend.text = element_text(size = 7)) +
  scale_fill_paletteer_d("beyonce::X11")

plot_grid(hpvq1male1, hpvq1male2, label_size = 5)

levels(nhanes_hpv_eligible$AnyHPVShots)
levels(nhanes2013$AnyHPVShots)
levels(nhanes2021$AnyHPVShots)

# Checked levels to confirm "Yes" answer to HPV vaccination history is the independent outcome being analyzed.

summary((glm(AnyHPVShots ~ Gender, data = nhanes2013, family = binomial())))
summary((glm(AnyHPVShots ~ Gender, data = nhanes2015, family = binomial())))
summary((glm(AnyHPVShots ~ Gender, data = nhanes2017, family = binomial())))
summary((glm(AnyHPVShots ~ Gender, data = nhanes2021, family = binomial())))

```

*In reviewing the results above...*

### Question #2: Age at First HPV Vaccination

```{r}

unknownfirstdoseage <- filter(nhanes_hpv_eligible, FirstDoseAge == "999")
knownfirstdoseage <- filter(nhanes_hpv_eligible, FirstDoseAge != "999" & !is.na(FirstDoseAge))

count(unknownfirstdoseage, name = 'Unknown First Dose Age')
count(knownfirstdoseage, name = 'Known First Dose Age')

q2summary <- nhanes_hpv_eligible |>
  select(DataCycle, Gender, FirstDoseAge) |>
  tbl_summary(by = DataCycle) |>
  remove_row_type(variables = c(FirstDoseAge), 
                  type = "missing") |>
  modify_caption("**Question #2: How old were you when you received your first dose of HPV vaccine?**") |>
  bold_labels()
q2summary

ggplot(knownfirstdoseage, aes(DataCycle, FirstDoseAge)) +
  geom_violin() + 
  geom_jitter(aes(color = FirstDoseAge), alpha = 0.5) +
  facet_grid(. ~ Gender) + 
  scale_color_viridis(option="plasma", limits=c(1,65)) +
  ylim(0, 60) +
  labs(title = "Age at First HPV Vaccination") +
  labs(x = "NHANES Data Cycle", y = "Age in years") + 
  theme(title = element_text(size = 10), 
        axis.text.x = element_text(angle = 45, size = 7),
        legend.title = element_text(size = 7), 
        legend.text = element_text(size = 7))

knownfirstdoseage |>
  group_by(Gender, DataCycle) |>
  summarize(AvgFirstHPVAge = mean(FirstDoseAge))
```

*Review results.*

### Question #3: Total Number of HPV Vaccines Received

```{r}

```

*Describe your results and include relevant tables, plots, and code/comments used to obtain them. You may refer to the @sec-methods as needed. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.*

## *Conclusion*

*This the conclusion. The @sec-results can be invoked here.*

## *Acknowledgements*

*Acknowledgements here.*

## References

1.  DeKloe, J., Urdang, Z. D., Martinez Outschoorn, U. E., & Curry, J. M. (2024). Effects of HPV vaccination on the development of HPV-related cancers: A retrospective analysis of a United States-based cohort.

2.  Centers for Disease Control and Prevention. *Quadrivalent Human Papillomavirus Vaccine: Recommendations of the Advisory Committee on Immunization Practices (ACIP)*. MMWR Early Release 2007;56(RR-12): 1–23.

3.  Centers for Disease Control and Prevention. *FDA Licensure of Quadrivalent Human Papillomavirus Vaccine (HPV4, Gardasil) for Use in Males and Guidance from the Advisory Committee on Immunization Practices (ACIP)*.  MMWR Morbidity and Mortality Weekly Report 2010;59(20): 630–632.

4.  Meites E, Kempe A, Markowitz LE. *Use of a 2-Dose Schedule for Human Papillomavirus Vaccination — Updated Recommendations of the Advisory Committee on Immunization Practices*. MMWR Morbidity and Mortality Weekly Report 2016;65: 1405–1408.

5.  Meites E, Szilagyi PG, Chesson HW, Unger ER, Romero JR, Markowitz LE. *Human Papillomavirus Vaccination for Adults: Updated Recommendations of the Advisory Committee on Immunization Practices*. MMWR Morbidity and Mortality Weekly Report 2019;68: 698–702.

6.  Ellingson, M. K., Sheikha, H., Nyhan, K., Oliveira, C. R., & Niccolai, L. M. (2023). Human papillomavirus vaccine effectiveness by age at vaccination: A systematic review. *Human vaccines & immunotherapeutics*, 19(2), 2239085.

7.  Centers for Disease Control and Prevention. (2024). *Brief overview of sample design, nonresponse bias assessment, and analytic guidelines for NHANES August 2021-August 2023*. Centers for Disease Control and Prevention. https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/overviewbrief.aspx?Cycle=2021-2023

8.  Centers for Disease Control and Prevention. (2024). *SAS Viewer*. Centers for Disease Control and Prevention. https://wwwn.cdc.gov/nchs/nhanes/SasViewer.aspx
